One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-05-08 doc/genesetdp.tex
text

\begin{figure}[htb]
  \begin{center}
		  \begin{tabular}[t]{c}
				\includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png} \\
				\includegraphics[width=0.5\textwidth]{figures/calibration_legend.png}
		\end{tabular}
  \end{center}
  \caption{{\bf Calibration of GeneSetDP and GeneSetMC.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway. To test the $p$~values uniformity we plotted them against their {\em normalized rank}, i.e. each $p$~value's (<rank>-0.5)/<total number of $p$~values>. We also added the calibration curve of BinoX, evaluated on the same random sets. The dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$, for comparison§§.}
  \label{fig:calibration}
\end{figure}

@@

Both our implementations, GeneSetDP and GeneSetMC, are intended for measuring enrichment of links between the query and the pathway set of genes. Most other NEA methods also report depletion of such links. One could easily modify our code to instead of only measuring the higher scoring tail measuring both tails of $N(s)$. However, this might result in a drop in sensitivity, so we have so far no implemented such a mechanism.

Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''. Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.

>>>>

\begin{figure}[htb]
  \begin{center}
		  \begin{tabular}[t]{c}
				\includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png} \\
				\includegraphics[width=0.5\textwidth]{figures/calibration_legend.png}
		\end{tabular}
  \end{center}
  \caption{{\bf Calibration of GeneSetDP and GeneSetMC.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway. To test the $p$~values uniformity we plotted them against their {\em normalized rank}, i.e. each $p$~value's (<rank>-0.5)/<total number of $p$~values>. We also added the calibration curve of BinoX, evaluated on the same random sets. The dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$, for comparison.}
  \label{fig:calibration}
\end{figure}

@@

Both our implementations, GeneSetDP and GeneSetMC, are intended for measuring enrichment of links between the query and the pathway set of genes. Most other NEA methods also report depletion of such links. One could easily modify our code to, instead of only measuring the higher scoring tail, measuring both tails of $N(s)$. However, this might result in a drop in sensitivity, so we have so far not implemented such a mechanism.

We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''. Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
